Realizability-Preserving Two-Moment DG Method for Fermionic Transport \tnoteref{ldrd-support}

### 2018-02-02 Documents/M1/realizableFermionicM1/sections/realizability.tex
Reworked the model section
Reworking of section 3: realizability

\section{Moment Realizability and Closures for Fermion}
\label{sec:realizability}

Our goal is to design a numerical method for solving the system of equations given by Eq.~\eqref{eq:momentEquations}, which approximates Eq.~\eqref{eq:boltzmann}.  
To this end, we want to obtain solutions that preserve features of the original problem.

Since we only consider the angular dependence of the distribution function in this section, we simplify the notation by suppressing the $\vect{z}$ and $t$ dependence, and write $f(\omegaNu,\vect{z},t)=f(\omegaNu)$.

@@

\begin{define}
  The moments $\vect{\cM}=\big(\cJ,\vect{\cH}\big)^{T}$ are realizable if they are consistent with a distribution function satisfying $0\le f(\omegaNu) \le 1$.
\end{define}

We begin by stating the bounds on the moments by restating results from Theorem~7.1 in \cite{banachLarecki_2017} in the following Lemma.  
\begin{lemma}
  Let the distribution function be bounded by $0\le f(\omegaNu) \le1$ for all $\omegaNu\in\bbS^{2}$.  
  Then $0\le\cJ\le1$ and $\big(1-\cJ\big)\,\cJ-|\vect{\cH}|\ge0$. 
  \label{lem: MomentRealizable} 
\end{lemma}
\begin{proof}
  First, let $\mu=\cos\thetaNu$ and write
  \begin{equation}
    \cJ=\f{1}{2}\int_{\bbS^{1}}\mathfrak{f}(\mu)\,d\mu,
    \quad\text{where}\quad
    \mathfrak{f}(\mu)=\f{1}{2\pi}\int_{0}^{2\pi}f(\mu,\phiNu)\,d\phiNu.  
  \end{equation}  
  Since $f(\omegaNu)\in[0,1]$, we have $\mathfrak{f}(\mu)\in[0,1]$, which implies $\cJ\in[0,1]$.  
  Next we need to show that $|\vect{\cH}|\le(1-\cJ)\,\cJ$.  
  Let $\widehat{\vect{n}}\in\bbR^{3}$ (independent of $\omegaNu$) be an arbitrary unit vector, and write
  \begin{equation}
    \widehat{\vect{n}}\cdot\vect{\cH}
    =\int_{\bbS^{1}}\mathfrak{f}(\mu)\,\mu\,d\mu,
    \label{eq:NdotH}
  \end{equation}
  where we have aligned the polar axis of the spherical momentum space coordinate system with $\widehat{\vect{n}}$; i.e., $\widehat{\vect{n}}\cdot\vect{\ell}(\omega)=\mu$.  
  First, following \cite{banachLarecki_2017}, we can write the integral on the right-hand side of Eq.~\eqref{eq:NdotH} as
  \begin{equation*}
    (1-\cJ)\cJ - \f{1}{2}\int_{-1}^{1-2\cJ}(1-2\cJ-\mu)\,\mathfrak{f}(\mu)\,d\mu
    -\f{1}{2}\int_{1-2\cJ}^{1}(2\cJ-1+\mu)\,[1-\mathfrak{f}(\mu)]\,d\mu.  
  \end{equation*}
  Since $\mathfrak{f}\in[0,1]$, $(1-2\cJ-\mu)$ is nonnegative on the interval $\mu\in[-1,1-2\cJ]$, and $(2\cJ-1+\mu)$ is nonnegative on the interval $\mu\in[1-2\cJ,1]$, we have $\widehat{\vect{n}}\cdot\vect{\cH}\le(1-\cJ)\,\cJ$.  
  In a similar manner, we can also write the integral on the right-hand side of Eq.~\eqref{eq:NdotH} as
  \begin{equation*}
    -(1-\cJ)\cJ-\f{1}{2}\int_{-1}^{2\cJ-1}(1-2\cJ+\mu)\,[1-\mathfrak{f}(\mu)]\,d\mu
    -\f{1}{2}\int_{2\cJ-1}^{1}(2\cJ-1-\mu)\,\mathfrak{f}(\mu)\,d\mu.  
  \end{equation*}
  Then, since $\mathfrak{f}\in[0,1]$, $(1-2\cJ+\mu)$ is nonpositive on the interval $\mu\in[-1,2\cJ-1]$, and $(2\cJ-1-\mu)$ is nonpositive on the interval $\mu\in[2\cJ-1,1]$, we have $\widehat{\vect{n}}\cdot\vect{\cH}\ge-(1-\cJ)\,\cJ$.  
  We therefore have $-(1-\cJ)\,\cJ\le\widehat{\vect{n}}\cdot\vect{\cH}\le(1-\cJ)\,\cJ$.  
  Since this holds for any unit vector $\widehat{\vect{n}}\in\bbR^{3}$, we must have $|\vect{\cH}|\le(1-\cJ)\,\cJ$.  
\end{proof}

@@

\begin{lemma}
  The set $\cR$, defined in Eq.~\eqref{eq:realizableSet}, is convex.  
\end{lemma}
\begin{proof}
  A convex set $Q$ is a set has the following property: 
  \begin{equation*}
  \theta a + (1-\theta)b \in Q,~\text{for}~\forall a, b \in Q~\text{and}~0\leq\,\theta\,\leq1. 
  \end{equation*} 
  Let $\vect{\cM}_{a}=\big(\cJ_{a},\vect{\cH}_{a}\big)^{T}$ and $\vect{\cM}_{b}=\big(\cJ_{b},\vect{\cH}_{b}\big)^{T}$ to be two arbitrary elements of $\cR$.
  If $\cR$ is convex, we will have $\vect{\cM}_{ab} = \big(\cJ_{ab},\vect{\cH}_{ab}\big)^{T} = \theta \vect{\cM}_{a} + (1-\theta)\vect{\cM}_{b}$ in $\cR$ with $0\leq\theta\leq1$.
  The first component of $\vect{\cM}_{ab}$ is
  \begin{equation*}
  \cJ_{ab} = \theta \cJ_{a} + (1-\theta)\cJ_{b}.
  \end{equation*}
  Since $\theta$, $\cJ_{a}$, $1-\theta$ and $\cJ_{b}$ are all nonnegative, $\cJ_{ab}$ is also nonnegative.
  Meanwhile, since $\cJ_{a}\leq1$ and $\cJ_{b}\leq1$, $\cJ_{ab}\leq1$.
  That's $\cJ_{ab} \in [0,1]$.
  Since the second derivative of $\gamma(\vect{\cM})$ is $-2$ which is always negative, $\gamma(\vect{\cM})$ is a concave function.
  Using Jensen's inequality, we have
  \begin{equation*}
  \gamma(\vect{\cM_{ab}}) \geq \theta\gamma(\vect{\cM_{a}}) + (1-\theta)\gamma(\vect{\cM_{b}}) \ge0.
  \end{equation*}
\end{proof}

The Fig.~\ref{fig:RealizableSetFermionic} illustrates how the convex realizable set $\cR$ looks like in $|\vect{\cH}| - \cJ$ coordinates.
Total 1000,000 distribution functions having form $\mathfrak{f}(\mu;a,b) = \dfrac{1}{e^{a+b\mu}+1}$ were evaluated.
The parameters, $a$ and $b$, were chosen from $[-100,100]$ independently.
The variable $\mu$ ranges from $-1$ to $1$ with $d\mu=\f{1}{128}$.
And the integrals in angular moment expressions are calculated using Trapezoid rule.
\begin{figure}
\centering
\includegraphics[width=1.0\linewidth]{figures/RealizableSetFermionic}
\caption[Numerical plot of the convex realizable set $\cR$.]{
A numerical plot of the convex realizable set $\cR$.
The black line are the boundary given by $\gamma(\vect{\cM}) = 0$. 
The red line are the boundary given by $|\vect{\cH}| = \cJ$.
Every light-blue dot represents a sample whose distribution function $\mathfrak{f}(\mu)$ is an arbitrary Fermi-Dirac function and its angular moments were given by integrals using Trapezoid rule.
Total 1000,000 $\mathfrak{f}(\mu)$ were sampled for making this plot.}
\label{fig:RealizableSetFermionic}
\end{figure}

\begin{lemma}
  Let $\big\{\cJ_{a},\vect{\cH}_{a},\vect{\cK}_{a}\big\}$ be defined as in Eq.~\eqref{eq:angularMoments} with distribution function $f_{a}\in[0,1]$.  
  Similarly, $\big\{\cJ_{b},\vect{\cH}_{b},\vect{\cK}_{b}\big\}$ be moments of a distribution function $f_{b}\in[0,1]$.  
  Let $\Phi^{\pm}(\vect{\cM})=\f{1}{2}\big(\vect{\cM}\pm\widehat{\vect{e}}\cdot\vect{\cF}(\vect{\cM})\big)$, where $\widehat{\vect{e}}\in\bbR^{3}$ is an arbitrary unit vector and $\vect{\cF}(\vect{\cM})=\big(\vect{\cH},\vect{\cK}\big)^{T}$.  
  Then
  \begin{equation}
    \vect{\cM}_{ab} \equiv \Phi^{+}(\vect{\cM}_{a})+\Phi^{-}(\vect{\cM}_{b})\in\cR.
  \end{equation}
\end{lemma}
\begin{proof}
  Using the same treatment we made in proof of Lemma~\ref{lem: MomentRealizable}, we can rewrite the first component of $\vect{\cM}_{ab}$ as
  \begin{align*}
  \f{1}{2}\big(\vect{\cJ_{a}} + \widehat{\vect{e}}\cdot\vect{\cH_{a}}+\vect{\cJ_{b}} - \widehat{\vect{e}}\cdot\vect{\cH_{b}}\big) = &\f{1}{2} \left\lbrace \f{1}{2}\int_{\bbS^{1}}\left[ \mathfrak{f}_{a}(\mu)\,(1+\mu) + \mathfrak{f}_{b}(\mu)\,(1-\mu)\right] \,d\mu \right\rbrace \\
  \equiv  &\f{1}{2}\int_{\bbS^{1}} \mathfrak{f}_{ab}(\mu)\,d\mu
  \end{align*}
  where $\mathfrak{f}_{ab}(\mu) \equiv \f{1}{2}\left[ \mathfrak{f}_{a}(\mu)\,(1+\mu) + \mathfrak{f}_{b}(\mu)\,(1-\mu)\right]$ and $\mathfrak{f}_{ab}(\mu)\in[0,1]$.
  And the second component is
  \begin{align*} 
   \f{1}{2}\big(\vect{\cH_{a}} + \widehat{\vect{e}}\cdot\vect{\cK_{a}}+\vect{\cH_{b}} - \widehat{\vect{e}}\cdot\vect{\cK_{b}}\big) =
   &\f{1}{2} \left\lbrace \f{1}{2}\int_{\bbS^{1}}\left[ \mathfrak{f}_{a}(\mu)\,(1+\mu) + \mathfrak{f}_{b}(\mu)\,(1-\mu)\right]\mu \,d\mu \right\rbrace \\
   \equiv  &\f{1}{2}\int_{\bbS^{1}} \mathfrak{f}_{ab}(\mu)\,\mu d\mu. 
  \end{align*}
  Applying Lemma~\ref{lem: MomentRealizable}, we know that $\vect{\cM}_{ab}\in\cR$.
\end{proof}

\subsection{Moment Closures}

@@

>>>>

\section{Moment Realizability and Algebraic Closures for the Fermionic Two-Moment Model}
\label{sec:realizability}

Our goal is to simulate a gas of massless fermions (i.e., classical neutrinos) and study their interactions with matter.  
The principal objective is to obtain the fermionic distribution function $f$ (or moments of $f$ as in the two-moment model employed here).  
The Pauli exclusion principle requires the distribution function to satisfy the condition $0 \le f \le 1$, which puts restrictions on the admissible values for the moments of $f$.  
In this paper, we seek to design a numerical method for solving the system of moment equations given by Eq.~\eqref{eq:momentEquations}, which preserves realizability of the moments; i.e., the moments evolve within the set of admissible values as dictated by Pauli's exclusion principle.  
(Since are concerned with the angular dependence of $f$ in this section, we simplify notation by suppressing the $\vect{z}$ and $t$ dependence and write $f(\omegaNu,\vect{z},t)=f(\omegaNu)$.)

@@

We begin with the following definition of moment realizability.  
\begin{define}
  The moments $\vect{\cM}=\big(\cJ,\vect{\cH}\big)^{T}$ are realizable if they can be obtained from a distribution function satisfying $0\le f(\omegaNu) \le 1~\forall~\omegaNu\in\bbS^{2}$.
\end{define}

Obviously, realizable moments must satisfy the bounds imposed by $f$.  
We proceed by stating the bounds on the moments by restating (with slightly modified proof) results from Theorem~7.1 in \cite{banachLarecki_2017} in the following Lemma.  
\begin{lemma}
  Let the distribution function be bounded by $0\le f(\omegaNu) \le1~\forall~\omegaNu\in\bbS^{2}$.  
  Then the moments $\vect{\cM}=(\cJ,\vect{\cH})^{T}$, defined as in Eq.~\eqref{eq:angularMoments}, satisfy $0\le\cJ\le1$ and $\big(1-\cJ\big)\,\cJ-|\vect{\cH}|\ge0$. 
  \label{lem:MomentRealizable} 
\end{lemma}
\begin{proof}
  First, let $\mu=\cos\thetaNu$ and $\bbS^{1}:=\{\,\mu\,|\,\mu\in[-1,1]\,\}$, and write
  \begin{equation}
    \cJ=\f{1}{2}\int_{\bbS^{1}}\mathfrak{f}(\mu)\,d\mu,
    \quad\text{where}\quad
    \mathfrak{f}(\mu)=\f{1}{2\pi}\int_{0}^{2\pi}f(\mu,\phiNu)\,d\phiNu.  
  \end{equation}  
  Since $f(\omegaNu)\in[0,1]$, we have $\mathfrak{f}(\mu)\in[0,1]$, which implies $\cJ\in[0,1]$.  
  Next we need to show that $|\vect{\cH}|\le(1-\cJ)\,\cJ$.  
  Let $\widehat{\vect{n}}\in\bbR^{3}$ (independent of $\omegaNu$) be an arbitrary unit vector, and write
  \begin{equation}
    \widehat{\vect{n}}\cdot\vect{\cH}
    =\int_{\bbS^{1}}\mathfrak{f}(\mu)\,\mu\,d\mu,
    \label{eq:NdotH}
  \end{equation}
  where we have aligned the polar axis of the spherical momentum space coordinate system with $\widehat{\vect{n}}$; i.e., $\widehat{\vect{n}}\cdot\vect{\ell}(\omega)=\mu$.  
  First, following \cite{banachLarecki_2017}, we can write the integral on the right-hand side of Eq.~\eqref{eq:NdotH} as
  \begin{equation*}
    (1-\cJ)\cJ - \f{1}{2}\int_{-1}^{1-2\cJ}(1-2\cJ-\mu)\,\mathfrak{f}(\mu)\,d\mu
    -\f{1}{2}\int_{1-2\cJ}^{1}(2\cJ-1+\mu)\,[1-\mathfrak{f}(\mu)]\,d\mu.  
  \end{equation*}
  Since $\mathfrak{f}\in[0,1]$, $(1-2\cJ-\mu)$ is nonnegative on the interval $\mu\in[-1,1-2\cJ]$, and $(2\cJ-1+\mu)$ is nonnegative on the interval $\mu\in[1-2\cJ,1]$, we have $\widehat{\vect{n}}\cdot\vect{\cH}\le(1-\cJ)\,\cJ$.  
  In a similar manner, we can also write the integral on the right-hand side of Eq.~\eqref{eq:NdotH} as
  \begin{equation*}
    -(1-\cJ)\cJ-\f{1}{2}\int_{-1}^{2\cJ-1}(1-2\cJ+\mu)\,[1-\mathfrak{f}(\mu)]\,d\mu
    -\f{1}{2}\int_{2\cJ-1}^{1}(2\cJ-1-\mu)\,\mathfrak{f}(\mu)\,d\mu.  
  \end{equation*}
  Then, since $\mathfrak{f}\in[0,1]$, $(1-2\cJ+\mu)$ is nonpositive on the interval $\mu\in[-1,2\cJ-1]$, and $(2\cJ-1-\mu)$ is nonpositive on the interval $\mu\in[2\cJ-1,1]$, we have $\widehat{\vect{n}}\cdot\vect{\cH}\ge-(1-\cJ)\,\cJ$.  
  We therefore have $-(1-\cJ)\,\cJ\le\widehat{\vect{n}}\cdot\vect{\cH}\le(1-\cJ)\,\cJ$.  
  Since this holds for any unit vector $\widehat{\vect{n}}\in\bbR^{3}$, we must have $|\vect{\cH}|\le(1-\cJ)\,\cJ$.  
\end{proof}

@@

\begin{lemma}
  The set $\cR$, defined in Eq.~\eqref{eq:realizableSet}, is convex.  
\end{lemma}
\begin{proof}
  In order to prove that $\cR$ is convex, we must show that a convex combination of elements in $\cR$ also belongs to $\cR$.  
  Let $\vect{\cM}_{a}=\big(\cJ_{a},\vect{\cH}_{a}\big)^{T}$ and $\vect{\cM}_{b}=\big(\cJ_{b},\vect{\cH}_{b}\big)^{T}$ to be two arbitrary elements in $\cR$.  
  Let $\vect{\cM}_{ab} = \theta\,\vect{\cM}_{a} + (1-\theta)\,\vect{\cM}_{b}$, with $0\leq\theta\leq1$.
  The first component of $\vect{\cM}_{ab}$ is
  \begin{equation*}
    \cJ_{ab} = \theta\,\cJ_{a} + (1-\theta)\,\cJ_{b}.
  \end{equation*}
  Since $\theta,\cJ_{a},\cJ_{b}\in[0,1]$, we have $\cJ_{ab} \in [0,1]$.  
  It is straightforward to verify that $\gamma(\vect{\cM})$ is a concave function of $\bcM$ if $\cJ\in[0,1]$.
  Jensen's inequality implies
  \begin{equation*}
  \gamma(\vect{\cM}_{ab}) \geq \theta\,\gamma(\vect{\cM}_{a}) + (1-\theta)\,\gamma(\vect{\cM}_{b}) \ge0.
  \end{equation*}
\end{proof}

Figure~\ref{fig:RealizableSetFermionic} illustrates the geometry of the convex set $\cR$ in $|\vect{\cH}| - \cJ$ coordinates (light blue region).
To make this plot, a total of $10^{6}$ distribution functions of the form $\mathfrak{f}(\mu;a,b) = \dfrac{1}{e^{a+b\mu}+1}$ ($\mu\in[-1,1]$) were generated, where the parameters $a$ and $b$ were chosen randomly and independently in the range $a,b\in[-10^{2},10^{2}]$.  
A uniform $\mu$-grid with $\Delta\mu=1/128$, was used to represent the distribution functions, and the angular moments were calculated using the trapezoid rule.  
\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{figures/RealizableSetFermionic}
  \caption{Illustration of the realizable set $\cR$ defined in Eq.~\eqref{eq:realizableSet}.  
  Each of the $10^{6}$ light-blue points in the figure represents a set of moments $\vect{\cM}=(\cJ,\vect{\cH})$, computed from a random Fermi-Dirac distribution (see text for details).  
  Black lines define the boundary of $\cR$; $\gamma(\vect{\cM}) = 0$.}
  \label{fig:RealizableSetFermionic}
\end{figure}

For the realizability-preserving scheme, we state some additional results.  
\begin{lemma}
  Let $\big\{\cJ_{a},\vect{\cH}_{a},\vect{\cK}_{a}\big\}$ be defined as in Eq.~\eqref{eq:angularMoments} with distribution function $f_{a}(\omegaNu)\in[0,1]\,\forall\,\omegaNu\in\bbS^{2}$.  
  Similarly, let $\big\{\cJ_{b},\vect{\cH}_{b},\vect{\cK}_{b}\big\}$ be moments of a distribution function $f_{b}\in[0,1]\,\forall\,\omegaNu\in\bbS^{2}$.  
  Let $\Phi^{\pm}(\vect{\cM})=\f{1}{2}\big(\vect{\cM}\pm\widehat{\vect{e}}\cdot\vect{\cF}(\vect{\cM})\big)$, where $\widehat{\vect{e}}\in\bbR^{3}$ is an arbitrary unit vector, and $\widehat{\vect{e}}\cdot\vect{\cF}(\vect{\cM})=\big(\widehat{\vect{e}}\cdot\vect{\cH},\widehat{\vect{e}}\cdot\vect{\cK}\big)^{T}$.  
  Then
  \begin{equation*}
    \vect{\cM}_{ab} = \big(\cJ_{ab},\vect{\cH}_{ab}\big)^{T} \equiv \Phi^{+}(\vect{\cM}_{a})+\Phi^{-}(\vect{\cM}_{b})\in\cR.
  \end{equation*}
\end{lemma}
\begin{proof}
  It is straightforward to rewrite the components of $\vect{\cM}_{ab}$ as
  \begin{equation*}
    \cJ_{ab}=\f{1}{4\pi}\int_{\bbS^{2}}f_{ab}(\omegaNu)\,d\omegaNu
    \quad\text{and}\quad
    \vect{\cH}_{ab}=\f{1}{4\pi}\int_{\bbS^{2}}f_{ab}(\omegaNu)\,\vect{\ell}(\omegaNu)\,d\omegaNu,
  \end{equation*}
  where $f_{ab}(\omegaNu)=\theta\,f_{a}+(1-\theta)\,f_{b}$ and $\theta(\omegaNu)=(1+\widehat{\vect{e}}\cdot\vect{\ell}(\omegaNu))/2\in[0,1]$.  
  Then, since $f_{ab}(\omegaNu)\in[0,1]\,\forall\,\omegaNu\in\bbS^{2}$, we have from Lemma~\ref{lem:MomentRealizable} that $\vect{\cM}_{ab}\in\cR$.  
\end{proof}

\subsection{Algebraic Moment Closures}

@@

\begin{figure}
  \centering
  \begin{tabular}{cc}
    \includegraphics[width=0.5\textwidth]{figures/MabWithBLME}
    \includegraphics[width=0.5\textwidth]{figures/MabWithCBME} \\
    \includegraphics[width=0.5\textwidth]{figures/MabWithBLKS}
    \includegraphics[width=0.5\textwidth]{figures/MabWithMI}
  \end{tabular}
  \caption{}
  \label{fig:MabWithDifferentClosure}
\end{figure}

%\begin{figure}[h!]
%  \begin{subfigure}[b]{0.5\textwidth}
%    \includegraphics[width=\textwidth]{figures/MabWithMI}
%    \caption{$\vect{\cM}_{ab}$ with Minerbo closure.}
%  \end{subfigure}
%  %
%  \begin{subfigure}[b]{0.5\textwidth}
%    \includegraphics[width=\textwidth]{figures/MabWithBLKS}
%    \caption{$\vect{\cM}_{ab}$ with Kershaw closure.}
%  \end{subfigure}
%  \begin{subfigure}[b]{0.5\textwidth}
%    \includegraphics[width=\textwidth]{figures/MabWithCBME}
%    \caption{$\vect{\cM}_{ab}$ with Chernohorsky \& Bludman maximum entropy closure.}
%  \end{subfigure}
%  \begin{subfigure}[b]{0.5\textwidth}
%    \includegraphics[width=\textwidth]{figures/MabWithBLME}
%    \caption{$\vect{\cM}_{ab}$ with Banach \& Larecki maximum entropy closure.}
%  \end{subfigure}
%  \label{fig:MabWithDifferentClosure}
%\end{figure}
